\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{algorithmic}

\title{CS 311 Assignment 1}
\author{Anton Rasmussen, collaborated with Jack McDonough}
\date{February 8, 2019}

\begin{document}

\maketitle

\section{Stable Matching Running Time:}
a) The Propose-and-Reject algorithm requires O(n) iterations for the following class of instances:\\ For any given college c in the instance, c's first choice is a student s such that s's first choice is c. With this instance there will be no rejections because every college will first propose to their favorite student and every student will first receive proposals from their top college. Each college will propose to and subsequently match with one and only one student, which implies the algorithm will run in linear time.\\\\
b) The Propose-and-Reject algorithm will require $\Omega(n^2)$ iterations for the following class of instances:\\ All n colleges have the exact same student preference order and all n students have the exact same college preference order. Then, regardless of the order in which the colleges pick, the algorithm will take $n(n + 1)/2$ iterations to resolve. This is because there exists a college $c_0$ such that $c_0$ is the least preferred among all the students. $c_0$ then, will be rejected n - 1 times for each college ahead of it while the second to last college $c_1$ will be rejected n - 2 times and so on. However, each college is also accepted in the end so we add an n to $\sum_{i = 1}^{i = n} n - i$ changing it to $\sum_{i = 0}^{i = n} n - i$. This evaluates to $n(n + 1)/2$. The function f(n) = $n(n + 1)/2$ is $\Omega(n^2)$ because there exists constants $c$ and $n_0$ such that $n(n + 1)/2 \geq cn^2$, namely $c = 1$ and $n_0 = 1$.

\section{Stable Matchings:}
a) The Gale-Shapley Algorithm is actually guaranteed to find a stable matching for any given instance of n students and n colleges. It's pseudo-code is as follows:
\begin{algorithmic}
\WHILE {some college is free}
    \STATE let c be any college that is free
    \STATE let s be the highest ranked student to whom c hasn't proposed
        \IF {s is free} 
            match c with s
        \ELSE
            \IF {$i\leq 3$}
                \STATE match c with s 
                \STATE unmatch c with s
            \ELSE
                \STATE s rejects c
        \ENDIF
\ENDWHILE
\end{algorithmic}
The intuition for this algorithm obviously came from what we learned the first lecture. As long as we're dealing with an instance involving 2 sets of n entities ranking each other, we proved that the Gale-Shapely algorithm is guaranteed to terminate and that it will always find a stable matching. The latter property is what we need to verify for this question and we can prove it by contradiction. Assume that there exists an instability involving student-college pairs c and s' along with c' and s. Assume that c prefers s over its current match s' and that c' prefers s' over its current match s. If c prefers s over s', c must have proposed to s before it was matched with s' because s is higher in c's preference list. If s accepted c's offer, it must have preferred c' over c because s chose to switch to c' when c' proposed. If s didn't accept c's offer, it must have been because s was already matched to a college c'' it preferred over c. Since students matches can only upgrade over the course of the algorithm, s must have preferred c' over c'' which implies s preferred c' over c. This contradicts our assumption. In Question 1 part b) it was determined and proved that the Propose-and-Reject algorithm (ie. the Gale-Shapley algorithm) requires $\Omega(n^2)$ iterations for some instances while it only requires $O(n)$ for others. This implies the algorithm isn't tightly bounded. The best way to describe the running time of the Gale-Shapley algorithm is to say it's $O(n^2)$. This is because each of the n colleges can propose to a maximum of n students and each attempted proposal causes the while loop to execute once. Once a college has proposed to a given student, they aren't allowed to propose to that student again, guaranteeing that the algorithm will terminate in at most $n^2$ iterations.\\\\
b) An example of an instance that has a weak instability is any instance in which the first college to propose (we'll call it $c$) is indifferent about two students both of whom prefer c over any other college. c will match with one of these students which will call student a, leaving student b to match with a different college. In this case b will prefer c over its current match while c will be indifferent between b and its current match. Hence, this instance is guaranteed to have a weak instability. 

\section{Big-O:}
a) \[f(n) = (1/2)n^2\] The smallest integer H such that f(n) is $O(n^H)$ is 2.\\ Proof: f(n) is $O(n^2)$ because for constants $c = 1/2$ and $n_0 = 0$, the following inequality holds for all $n \geq n_0$: $(1/2)n^2 \leq cn^2$.\\

The largest positive real constant L such that f(n) is $\Omega(n^L)$ is 2.\\ Proof: f(n) is $\Omega(n^2)$ because for constants $c = 1/2$ and $n_0 = 0$, the following inequality holds for all $n \geq n_0$: $(1/2)n^2 \geq cn^2$. To show that this is the largest positive real constant, consider the inequality f(n) $\geq n^{(2 + \epsilon)}$. Substitute in f(n) and use exponent rules to get $(1/2)n^2 \geq n^2n^{\epsilon}$. Cancel an $n^2$ from both sides and it's clear that a constant doesn't grow faster than a polynomial for any $n \geq n_0$.\\

b) \[f(n) = n(log(n))^3\] The smallest integer H such that f(n) is $O(n^H)$ is 2.\\ Proof: f(n) is $O(n^2)$ because for constants $c = 5$ and $n_0 = 0$, the following inequality holds for all $n \geq n_0$: $n(log(n))^3 \leq cn^2$.\\

The largest positive real constant L such that f(n) is $\Omega(n^L)$ is 1.\\ Proof: f(n) is $\Omega(n)$ because for any constants $c = 1$ and $n_0 = 2$, the following inequality holds for all $n \geq n_0$: $n(log(n))^3 \geq cn$. To show that this is the largest positive real constant, consider the inequality f(n) $\geq n^{(1 + \epsilon)}$. Substitute in f(n) and use exponent rules to get $n(log(n))^3 \geq nn^{\epsilon}$. Cancel an $n$ from both sides and it's clear that a logarithm doesn't grow faster than a polynomial for any $n \geq n_0$.\\

c) \[f(n) = \sum_{i=1}^{log(n)} n/(2^i)\] The presence of the logarithm in the summation admittedly made it difficult to simplify the formula algebraically. Instead I derived f(n) by checking the value of the formula for powers of 2. I found that the result for a given n where n was a power of 2 was $\sum_{i=0}^{n - 1} 2^n$. This in turn resulted in the formula f(n) = n - 1 which is proved by induction://
Let P(n) be $\sum_{i=1}^{log(n)} n/(2^i) = n - 1$\\
Base Case: Since log(1) = 0, the summation will equal 0 which equals $1 - 1 = 0$\\
Inductive Step: $(n(n + 1)/2)^2 + (n + 1)^3$ = $(n^4+2n^3+n^2)/4 + (n + 1)^3$ = $(n^4+6n^3+13n^2+12n+4)/4$ = $((n+1)(n+2)/2)^2$.\\
The smallest integer H such that f(n) is $O(n^H)$ is 1.\\ Proof: f(n) is $O(n)$ because for constants $c = 1$ and $n_0 = 0$, the following inequality holds for all $n \geq n_0$: $n - 1 \leq cn$.\\

The largest positive real constant L such that f(n) is $\Omega(n^L)$ is also 1.\\ Proof: f(n) is $\Omega(n)$ because for constants $c = 1/2$ and $n_0 = 2$, the following inequality holds for all $n \geq n_0$: $n - 1 \geq cn$. To show that this is the largest positive real constant, consider the inequality f(n) $\geq n^{(1 + \epsilon)}$. Substitute in f(n) and use exponent rules to get $n - 1 \geq nn^{\epsilon}$. Cancel an $n$ from both sides and it's clear that a polynomial of a negative degree doesn't grow faster than a polynomial of a positive degree for any $n \geq n_0$ regardless of the value of $n_0$.\\\\

d) \[f(n) = \sum_{i=1}^{n} i^3\] The smallest integer H such that f(n) is $O(n^H)$ is 4.\\ Proof: The first step is to write f(n) as a function of n. The intuition for this formula comes from the fact that sum of the first n cubes equals $k^2$, where k is the nth triangle number. Knowing the formula for triangle numbers already, one can simply square that formula to find that f(n) = $(n(n + 1)/2)^2$ which forms a quartic polynomial when simplified.\\
The largest positive real constant L such that f(n) is $\Omega(n^L)$ is 4.\\ Proof: As shown in the previous proof, f(n) is a quartic polynomial and so it should be tightly bounded. f(n) $\Omega(n^4)$ because for constants $c = 1/4$ and $n_0 = 0$, the following inequality holds for all $n \geq n_0$: $(n(n + 1)/2)^2 \geq cn^4$. To show that this is the largest positive real constant, consider the inequality f(n) $\geq n^{(4 + \epsilon)}$. Substitute in f(n) and divide both sides by $n^4$ to get $1/4(1 + 2/n + 1/n^2) \geq n^{\epsilon}$. A polynomial of a negative degree clearly doesn't grow faster than a polynomial of a positive degree so we have a contradiction.\\\\

e) \[f(n) = 2^{log(n)^2}\] There doesn't actually exist an integer H such that f(n) is $O(n^H)$.\\ Proof: For there to exist such an integer, the inequality $2^{log(n)^2} \leq cn^H$ must hold beyond some threshold $n_0$ some constant $c$, and some integer H. $2^{log(n)^2}$ simplifies to $2^{log(n)*log(n)}$ which equals $2^{(log^n)^{log(n)}}$ which equals $n^{log(n)}$. Taking the nth log of both sides of the inequality yields $log(n) \leq H + log_n(c)$. A constant clearly doesn't grow faster than a logarithm function for $n \geq n_0$ regardless of the value of $n_0$.\\

The largest positive real constant L such that f(n) is $\Omega(n^L)$ is undefined for the same reason as the previous proof. f(n) will be $\Omega(n^L)$ for any value of L because when the inequality is simplified, a logarithm function will still grow faster than a constant.\\\\

\section{Asymptotics:}
a) The running time of the algorithm is $O(n^3)$.\\
Explanation: The first loop in the algorithm clearly executes n times. With each execution of the first loop, the second loop will execute n - 1, n - 2, ... , n - n + 1, 0 times. Adding these quantities together yields the expression $(n^2 + n)/2$ which equals $n(n-1)/2$. This represents the total number of times the second loop executes. However, the operation of computing the maximum of entries A[i] to A[j] requires a traversal of A that iterates j - i times. In the worst case, j - i equals the length of A minus 1 so the algorithm's runtime is no better than the function $f(n) = n^2(n-1)/2$. f(n) is $O(n^3)$ because for all $n \geq n_0$, $f(n) \leq cn^3$ for $c = 1$ and $n_0 = 0$.\\
b) The running time of the algorithm is also $\Omega(n^3)$.\\
Explanation: $f(n) = n^2(n-1)/2$ is $\Omega(n^3)$ because for all $n \geq n_0$, $f(n) \geq cn^3$ for all $c = 1/4$ and $n_0 = 5/3$.\\
c) Faster algorithm:\\
\begin{algorithmic}
\FORALL{i = 0, 1, 2, ..., n}
    \STATE compute the maximum of the entries A[0] - A[i]
    \STATE store maximum value in B[0][i]
\ENDFOR
\FORALL{i = 1, 2, ..., n}
    \FORALL{j = i + 1, ..., n}
        \STATE Store the value at B[0][j] in B[i][j]
    \ENDFOR
\ENDFOR
\end{algorithmic}\\
The intuition for this algorithm came from the fact that any column in the matrix B will have the same entries as long as $j > i$. This is because the top-most entry in a given column determines the entries below it. The algorithm takes advantage of this property by iterating through A only n times (for each column in B). Then the algorithm fills in the remaining spaces in B by referencing the entry in the first row of the each of column (constant time operation).\\
Proof: Let's consider entry c located at B[0][j]. Then c, by definition, is the maximum of the entries in A[0] - A[j]. If c is the maximum of the entries in this range, c must also be the maximum of the entries in A[i] - A[j] for $i > 0$ because these ranges are smaller. Since the entry in B[i][j] for $i > 0$ is determined by computing the maximum of these smaller ranges, these positions in B must be occupied by c.\\
This algorithm runs in time $O(n^2)$. The first loop executes only n times but the second set of loops are nested. As in the original algorithm, for each time the first loop executes, the second loop will execute n - 1, n - 2, ... , 1 times. Adding these quantities together yields the expression $(n^2 + n)/2$ which equals $n(n-1)/2$. The function $g(n) = n(n-1)/2$ is $O(n^2)$ because for all $n \geq n_0$, $g(n) \leq cn^2$ for $c = 1$ and $n_0 = 0$. $O(n^2)$ is a better runtime than $O(n^3)$ and the $\lim_{n\to\infinty} f(n)/g(n) = 0$.\\

\section{DFS and BFS:}
Suppose we have a connected graph G = (V, E)
and a vertex u ∈ V . If we run DFS from u, we obtain a tree T. Suppose that if we run BFS from u we
obtain exactly the same tree T. Prove that G = T.\\

(3.4) Let T be a breadth-first search tree, let x and y be nodes in T belonging
to layers Li and Lj respectively, and let (x, y) be an edge of G. Then i and j differ
by at most 1.\\
(3.7) Let T be a depth-first search tree, let x and y be nodes in T, and let (x, y) be an edge of G that is not an edge of T. Then one of x or y is an ancestor
of the other.\\
Proof by Contradiction: Keeping in mind that T is the result of both a breadth-first search and a depth-first search of G from the vertex u, we assume that there exists an edge (x, y) in G that isn't in T. Then by Fact (3.7), one of x or y is an ancestor of the other in T (1). T is also a breadth-first search tree and we know that x and y are nodes in T and that (x, y) is an edge in G from our initial assumption. Then by Fact (3.4) we know that x and y belong to layers Li and Lj in T where i and j differ by at most 1 (2). By (1) we know that x and y are in the same branch of T and since x and y are distinct nodes, we know by (2) that the layers in which x and y reside differ by exactly. The only way that both of these properties are true is if there is and edge from x to y in T. This contradicts our initial assumption and completes the proof.  

\section{Butterfly ID:}
Algorithm to determine consistency of judgements:\\
\begin{itemize}
  \item The collection of m judgements is actually a representation of the graph where each butterfly pair is an edge between two vertices and each judgement is a label on the edge
  \item Treating this as a graph, one can perform a breadth-first search on the graph starting at any arbitrary vertice
  \item While the elements are on the open list, label each one either a or b (label the first vertice 'a')
  \item When a given vertice's neighbors are put onto the open list, they should be labeled the same as the vertice if the edge between them is labeled same and different if the edge between them is labeled different
  \item When the elements are taken off the open list they should be put in a collection visited where they are stored along with their labels
  \item Prior to the BFS search, create a counter which increments with each iteration of the BFS. When this counter reaches n where n is the number of distinct butterflies, elements shouldn't be put into visited when they are discovered. Instead they are compared to the elements already in visited
  \item If the element is the same butterfly as an element in visited but the two elements have different labels, then the judgements weren't consistent
  \item If the BFS terminates, and the above condition is false for the last element, then the judgements were indeed consistent
\end{itemize}

Note: It probably took me around 10-12 hours to do this assignment but I think a lot of it had to do with learning Latex for the first time as well as grappling with Problem 6.
\end{document}
